Soss

### 2020-01-02 paper/paper.tex
edits

A given inference algorithm can be expressed in terms of a collection of functions on a model or joint distribution. For example, we often need to evaluate the log-density or its gradient, map continuous variables into $\mathbb{R}^n$, or draw random samples. In Soss, these \emph{inference primitives} generate model-specific code at execution time.

@@

\begin{description}  
    \item[\texttt{rand}] draws a random point or sample from a given distribution.
    \item[\texttt{particles}] draws a \emph{systematic sample}, as described in \cite{Douc2005}.
    \item[\texttt{logpdf}] takes a distribution and a point, and returns the log-density at that point.
    \item[\texttt{xform}] takes a distribution and some known data, and returns a bijection from $\mathbb{R}^n$ to the space of unknowns. This is useful for algorithms like Hamiltonian Monte Carlo \cite{Neal2011}.
    \item[\texttt{symlogpdf}] 
    \item[\texttt{codegen}]  
\end{description}

@@

@@

\subsubsection*{Conditional predictive models}

\subsubsection*{Causal interventions (``Do'' operator)}

\subsubsection*{Markov blankets}

\begin{minted}{julia}
julia> markovBlanket(m, :α)
    @model (I, σI, β, σ, J) begin
            α ~ Normal(0, σI) |> iid(I)
            yhat = α .+ β'
            y ~ For(I, J) do i, j
                    Normal(yhat[i, j], σ)
                end
        end 
\end{minted}

\subsection{Symbolic Manipulation}

\subsubsection*{Symbolic Log-density}

\subsubsection*{Code Generation}

@@

\begin{minted}{julia}
  abstract type LCTerm end
  struct Lam{Arg, Body <: LCTerm} <: LCTerm end
  struct App{Fn<:LCTerm, Arg<:LCTerm} <: LCTerm end
  struct Var{N} <: LCTerm end

  cg(::Type{Lam{Arg, Term}}) where {Arg, Term} = 
    :($Arg -> $(cg(Term)))
  cg(::Type{App{Fn, Arg}}) where {Fn, Arg} = 
    :($(cg(Fn))($(cg(Arg))))
  cg(::Type{Var{N}}) where N = N

  function tolc(expr)
    @match expr begin
      :($f($arg)) => App{tolc(f), tolc(arg)}
      :($x => $y) => Lam{x, tolc(y)}
      a::Symbol => Var{a}
    end
  end

  @gg function lc(::Type{Term}) where Term <: LCTerm
      cg(Term)
  end

  macro lc(term)
    lc(tolc(term))
  end

  x = Core.Box()
  y = Core.Box()
  z = Core.Box()
  f(x) = (x, z)
  @assert (@lc x => x)(x) === (@lc y => y)(x)
  @assert (@lc f => x => f(x))(f)(y) === (y, z)
  \end{minted}

>>>>

A given inference algorithm can be expressed in terms of a collection of functions on a model or joint distribution. For example, we often need to evaluate the log-density or its gradient  or draw random samples. In Soss, these \emph{inference primitives} generate model-specific code at execution time.

@@

\begin{description}  
    \item[\texttt{rand}] draws a random point or sample from a given distribution.
    \item[\texttt{particles}] draws a \emph{systematic sample}, as described in \cite{Douc2005}.
    \item[\texttt{logpdf}] takes a distribution and a point, and returns the log-density at that point.
    \item[\texttt{xform}] takes a distribution and some known data, and returns a bijection from $\mathbb{R}^n$ to the space of unknowns. This is useful for algorithms like Hamiltonian Monte Carlo \cite{Neal2011}.
    \item[\texttt{symlogpdf}] takes a model, and returns a SymPy \cite{10.7717/peerj-cs.103} representation of the log-density, with some additional Julia-based simplifications.
    \item[\texttt{codegen}] calls \jl#symlogpdf#, and uses the result to generate a more efficient version of \jl#logpdf#.
\end{description}

@@

\subsection{Functions on Models}

arguments
sampled
assigned
parameters
variables

toposort
poset
digraph

@@

\begin{description}
    \item[\texttt{Do}] implements Pearl's {\footnotesize\texttt{do}} operator for causal intervention \cite{Pearl1995}.\footnote{Our {\footnotesize\texttt{Do}} operator is capitalized because {\footnotesize\texttt{do}} is a keyword in Julia.}
    \item[\texttt{predictive}] is like \jl#Do#, but only returns the strongly-connected components containing the specified variables. This is useful for sampling from the posterior predictive distribution.
    \item[\texttt{markovBlanket}]  takes a model and a variable, and returns a model representing the Markov blanket at the given variable.
\end{description}

@@

% \begin{minted}{julia}
%   abstract type LCTerm end
%   struct Lam{Arg, Body <: LCTerm} <: LCTerm end
%   struct App{Fn<:LCTerm, Arg<:LCTerm} <: LCTerm end
%   struct Var{N} <: LCTerm end

%   cg(::Type{Lam{Arg, Term}}) where {Arg, Term} = 
%     :($Arg -> $(cg(Term)))
%   cg(::Type{App{Fn, Arg}}) where {Fn, Arg} = 
%     :($(cg(Fn))($(cg(Arg))))
%   cg(::Type{Var{N}}) where N = N

%   function tolc(expr)
%     @match expr begin
%       :($f($arg)) => App{tolc(f), tolc(arg)}
%       :($x => $y) => Lam{x, tolc(y)}
%       a::Symbol => Var{a}
%     end
%   end

%   @gg function lc(::Type{Term}) where Term <: LCTerm
%       cg(Term)
%   end

%   macro lc(term)
%     lc(tolc(term))
%   end

%   x = Core.Box()
%   y = Core.Box()
%   z = Core.Box()
%   f(x) = (x, z)
%   @assert (@lc x => x)(x) === (@lc y => y)(x)
%   @assert (@lc f => x => f(x))(f)(y) === (y, z)
%   \end{minted}
